Tail associations and their impact on extinction risk

### 2019-11-06 MainText.Rmd
small edits on final read through

---
title: "Tail associations and their impact on extinction risk"
author: "Shyamolina Ghosh, Daniel Reuman"
date: ""
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

\noindent Extreme climatic events are becoming more frequent and more intense
due to climate change. Furthermore, there is reason to believe extreme 
climatic events may modify "tail associations" between values of an environmental 
variable measured in different locations, i.e., associations that 
occur between values in the left or right tails of the distributions 
of the variable that occur in the locations. 
Two positively associated variables can be principally "left-tail associated" (i.e., more 
correlated when they take low than high values) or "right-tail associated"
(more correlated when they take high than low values), even with the same
overall correlation coefficient in both cases. We tested whether the nature of 
spatial tail associations of environmental variables can influence metapopulation 
extinction risk. For instance, if low values of a variable reduce the growth 
rates of local
populations, one may expect that left-tail associations increase
metapopulation extinction risks because then environmental catastrophes are
spatially synchronized, presumably reducing the potential for rescue effects.
We also tested, in the context of non-spatial stage-structured matrix models, 
whether tail associations between stage-specific vital rates may
influence extinction risk. 
We showed that density dependence interacts with the
propensity for tail associations to influence metapopulation extinction risk: for 
population models
showing undercompensatory density dependence, left-tail associations in environmental 
variables typically accentuated
and right-tail associations mitigated extinction risk; whereas the 
reverse was usually true for models showing overcompensatory density dependence.
For the non-spatial, stage structured models we considered, left-tail associations 
between vital rates did accentuate extinction risk compared to right-tail
associations, but the effect was small. 
Tail associations and their asymmetries are taken into account in assessing risks
in finance and other fields, but to our knowledge our study is one of the first
to consider how tail associations influence population extinction risk. 
Our modelling results 
provide an initial demonstration of a new mechanism influencing extinction risks,
and, in our view, should help motivate more comprehensive
study of the mechanism and its importance for real populations in future work.

\noindent \textbf{\textit{Keywords:}} extinction risk, extreme events, tail association, density-dependent model, copula

@@

<!--ECEs have a lot of impacts, it is important to understand their impacts-->
Over the past few decades, extreme climatic events (ECEs) such as
temperature or precipitation extremes, heat-waves, and extreme wildfires
have become more common or more extreme due to anthropogenic climate change 
[@ummenhofer2017extreme]. ECEs are well known to produce severe
impacts at the population, community and ecosystem levels [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@felton2017integrating], and thus the goal of improving our understanding of
the variety of ecological impacts of ECEs is an important applied goal for 
modern ecology [@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling]. Frequent extreme events 
can help alter the mean temperature or precipitation of an area, 
can be associated with changes through time in the overall variance of 
climate signals, and can directly contribute to changes in the symmetry or the 
skewness of distributions of climate signals through time 
[@hansen2012perception]. All such changes have the potential to influence 
populations, including their risk of extirpation from an area 
[@GarciaCarerras2013; @Vasseuretal2014].
For instance, @frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs on European shags 
via their effects on temporal variability in 
juvenile, immature and adult survival rates.
@mcdermott2017sensitivity investigated the sensitivity 
of $41$ UK butterfly species to four different types of ECEs. 
At the community level, research has shown that ECEs can reorder
the dominance hierarchy of species in an area, sometimes leading to 
selective local extinctions, and can also produce effects that cascade 
across trophic levels [@thomas2004extinction; @edgar2010nino; @moreno2011extreme; @hoover2014resistance]. 
For example, @chiu2013impact showed that extreme floods affected 
the annual survival of a prey community, in turn influencing predators in 
aquatic as well as adjacent riparian habitats. 
@dreesen2014successive, in an experimental study, showed that 
ECEs can reduce the ability of plant communities to withstand subsequent ECEs
that occur after a short time interval.

@@

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
for both Q1 and Q2 instead of attempting mathematical 
analysis and/or using models that are more realistic, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms, outlined above, which suggest that
asymmetries of tail association may make a difference for extinction risks. 
By so doing we hope to open these potentially practically important questions to 
wider examination by other researchers. Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.

@@

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Spearman 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Spearman's $\rho$ was the same, $\rho =$ `r srho_common`,
up to sampling variance (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. 
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables.

<!--FIG : schematic diagram for partial spearman correlation, adapted from BIVAN-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=6 cm]{./Results/PartialSpearmanFig.pdf}
\caption{The partial Spearman correlation, $\cor_{l_b,u_b}(u, v)$, within a band can be computed
for any band to describe how the strength of association between $u$ and $v$
varies from one part of the two distributions to another. Diagonal lines 
show two bands, the data in the right/upper band showing
stronger association than those in the left/lower band. This figure was reproduced
largely following Fig. 7 of Ghosh \emph{et al.} (2019).
\label{PartialSpearmanFig}}
\end{center}
\end{figure*}

@@

<!--
We plotted extinction risks along the y-axis
for both left tail and right tail associated noises 
against the varying range of $r$ along x-axis when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal to avoid any rescue effect among patches. 
For each model with extreme (or moderate) left- and right-tail associated noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risks at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risks from both noises overlapped
with each other. This could give us an idea of how extinction risks affected by the model parameter 
$r$ that controlled the underlying dynamics of the model, i.e., undercompensatory 
or overcompensatory.
-->

@@

<!--Q1 : age-structured model results-->
For our age-structured, non-spatial model, extinction risks were higher when 
associations between the vital rates $s_e$ and $f_a$ were principally for low values
of these vital rates (left-tail association), and were lower when associations
were principally for higher values; but the effect was minor for this model (Fig. \ref{fig_extrisk_agestr}).
Though extinction risks were overall higher for the density-dependent version of our model compared
to density-independent version, extinction risk differences across tail association 
scenarios were similar for the two versions. Because the same Spearman correlation (up to sampling
variance) was used in all simulations, the differences between lines in the panels
of Fig. \ref{fig_extrisk_agestr} are attributable solely to the changes we made in the
asymmetry of tail associations, and answer Q1. Though differences were significant (confidence
bands in Fig. \ref{fig_extrisk_agestr}, which are very narrow, did not overlap),
they were also small, corresponding to only a few percentage points of additional extinction
risk after 25 years. A potential reason for this outcome is elaborated in the Discussion.
<!--discussion later: add Tuljapurkar's approximation stuff-->

<!--Q2 : metapopulation results-->
For our metapopulation simulations, for low values of $r$, left-tail
associations of environmental noises accentuated extinction risks, and right-tail associations
mitigated them; but the reverse was true for high values of $r$ in the range
$(r_{min},r_{bf})$. Thus 
density dependent models can either produce similar results, regarding the influence of 
tail associations on extinction risk, as the density-independent models of @Ghosh_copula, or
opposite results, depending on the nature of density dependence. 
For each model of Table \ref{tab_model_summary}, using a value of $r$ on
the low end of the range $(r_{min},r_{bf})$, extinction risk was highest
for extreme left-tail associated noise, lowest for extreme right-tail associated
noise, and was intermediate for the moderate tail association scenarios, for
both local (Fig. \ref{fig_extrisk_allmodel_localdisp}, left panels) and global 
(Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}, left panels) dispersal, and for 
essentially all the dispersal rates, $d$, that we considered. However, 
using a value of $r$ on the high end of the range $(r_{min},r_{bf})$,
extinction risks were ordered oppositely (right panels of Figs 
\ref{fig_extrisk_allmodel_localdisp} and \ref{SM-fig_extrisk_allmodel_globaldisp}). 
These results help address Q2 from the Introduction.
Intermediate values of $r$ are shown in Figs. \ref{SM-fig_ricker_risk_local_disp_scl_1} - 
\ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} for local dispersal and 
Figs. \ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6} for global dispersal among 
habitat patches.

@@

<!--summary results:-->
Our results combined with earlier results 
show that asymmetries in tail associations between environmental variables in
different locations can have a substantial influence on metapopulation extinction risk,
for density-dependent as well as for density-independent 
models. These results
extend the result of @Ghosh_copula, which was for a density-independent model. For a
density-independent model, @Ghosh_copula corroborated the intuition
that, when "harmful" environmental events 
(i.e., events which reduce the growth rates of local populations)
are correlated across space,
extinction risk of metapopulations will be accentuated compared 
to when "beneficial" events
(i.e., events which increase the growth of local populations)
are spatially correlated, even if overall correlation is the same in 
both scenarios. This was presumably because the first of these scenarios reduces the 
potential for rescue effects. For the density-dependent metapopulation 
models we considered, the same 
result held, typically, but not always, for undercompensatory model dynamics. However,
for overcompensatory dynamics, the oposite result held, again typically, but not always: 
when "harmful" events were correlated across space,
extinction risk of metapopulations was actually mitigated compared 
to when "beneficial" environmental events were 
spatially correlated by the same amount. This was probably because the correlated 
beneficial events caused populations to exceed carrying capacity, and 
to subsequently crash, across large areas. In contrast 
to the metapopulation case, our results 
suggest that asymmetry of tail associations between vital rates
in a stage-structured, non-spatial stochastic model may be of limited 
importance for extinction risk. In our stage-structured, non-spatial model, extinction 
risk was only weakly influenced by asymmetry of tail associations
between vital rates. Our results provide suggestive, rather
than definitive evidence, here, because we considered 
only one model (but see below).

<!--linking back to ECEs-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and even specific events can now be attributed to
human-induced climate change [@ummenhofer2017extreme]. We here provide a fuller 
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places, and thereby influence metapopulation
extinction risk through the mechanisms we explored in this paper. 
If $X_i(t)$ denotes the value of an environmental variable (e.g., a summer maximum 
temperature, or winter minimum temperature, spring rainfall, etc.) 
in location $i$ in year $t$,
@ummenhofer2017extreme provide a diagram of how ECEs may be increasing in frequency 
and intensity due to changes in the distribution of $X_i(t)$ such as changes
in the mean, variance, or skewness of this distribution. Although changes in 
the correlation between $X_i(t)$ and $X_j(t)$ are increasingly documented, 
and may also be a consequence of climate change (e.g., @Black2018),
we are unaware of any studies examining the extent to which tail associations
between $X_i(t)$ and $X_j(t)$, and asymmetries of tail association, may be 
changing. Documented changes in correlation between $X_i(t)$ and $X_j(t)$
may be due partly or wholly to changes in tail associations. If ECEs are becoming
more spatially extensive, as seems likely given that they are becoming
more intense [@ummenhofer2017extreme], then the conditional probability that both of
$X_i(t)$ and $X_j(t)$ are influenced by an ECE given that one of them is so
influenced should be increasing. This should contribute to increased association
of the two variables in one of their tails (which tail depends on the type of 
ECE and what the variable $X$ is). A study of changes through time in 
tail associations of weather variables would be straightforward and 
potentially important given the mechanisms we have explored.
<!--***DAN: Shyamolina, another aspect of an exploration you may do on the 
the geography of tail association may be to look at how tail association 
may be changing. This relates to ECEs, so may help better justify the work
to a funder. You had proposed to look at geography of tail association in 
NDVI, and that makes good sense. You could also look at how tail association
is changing in NDVI (the records are about 35 years long, so you could just
about do this). Even better (because data are better), you could get weather
station records (lots of these can be downloaded, over 1000 stations
have been operating for over a century continuously in North America) and 
look at how tail associations are changing in those data.-->

>>>>

---
title: "Tail associations and their impact on extinction risk"
author: "Shyamolina Ghosh, Daniel C. Reuman"
date: ""
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

\noindent Extreme climatic events are becoming more frequent and more intense
due to climate change. Furthermore, there is reason to believe extreme 
climatic events may modify "tail associations" between values of an environmental 
variable measured in different locations, i.e., associations that 
occur between values in the left or right tails of the distributions 
of the variable that occur in the locations. 
Two positively associated variables can be principally "left-tail associated" (i.e., more 
correlated when they take low values than when they take high values) or "right-tail associated"
(more correlated when they take high than low values), even with the same
overall correlation coefficient in both cases. We tested whether the nature of 
spatial tail associations of environmental variables can influence metapopulation 
extinction risk. For instance, if low values of an environmental variable reduce the growth 
rates of local
populations, one may expect that left-tail associations increase
metapopulation extinction risks because then environmental "catastrophes" are
spatially synchronized, presumably reducing the potential for rescue effects.
We also tested, in the context of non-spatial stage-structured matrix models, 
whether tail associations between stage-specific vital rates may
influence extinction risk. 
We showed that density dependence interacts with the
propensity for tail associations to influence metapopulation extinction risk: for 
population models
showing undercompensatory density dependence, left-tail associations in environmental 
variables typically accentuated
and right-tail associations mitigated extinction risk; whereas the 
reverse was usually true for models showing overcompensatory density dependence.
For the non-spatial, stage-structured models we considered, left-tail associations 
between vital rates did accentuate extinction risk compared to right-tail
associations, but the effect was small. 
Tail associations and their asymmetries are taken into account in assessing risks
in finance and other fields, but to our knowledge our study is one of the first
to consider how tail associations influence population extinction risk. 
Our modelling results 
provide an initial demonstration of a new mechanism influencing extinction risks,
and, in our view, should help motivate more comprehensive
study of the mechanism and its importance for real populations in future work.

\vspace{12pt}
\noindent \textbf{\textit{Keywords:}} extinction risk, extreme events, tail association, density-dependent model, copula

@@

<!--ECEs have a lot of impacts, it is important to understand their impacts-->
Over the past few decades, extreme climatic events (ECEs) such as
temperature or precipitation extremes, heat-waves, and extreme wildfires
have become more common or more extreme due to anthropogenic climate change 
[@ummenhofer2017extreme]. ECEs are well known to produce severe
impacts at the population, community and ecosystem levels [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@felton2017integrating], and thus the goal of improving our understanding of
the variety of ecological impacts of ECEs is an important applied goal for 
modern ecology [@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling]. Frequent extreme events 
can help alter the mean temperature or precipitation of an area, 
can be associated with changes through time in the overall variance of 
climate signals, and can directly contribute to changes in the symmetry or the 
skewness of distributions of climate signals through time 
[@hansen2012perception]. All such changes have the potential to influence 
populations, including through risks of species extinction or extirpation from an area 
[@GarciaCarerras2013; @Vasseuretal2014].
For instance, @frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs on European shags 
via their effects on temporal variability in 
juvenile, immature and adult survival rates.
@mcdermott2017sensitivity investigated the sensitivity 
of $41$ UK butterfly species to four different types of ECEs. 
At the community level, research has shown that ECEs can reorder
the dominance hierarchy of species in an area, sometimes leading to 
selective local extinctions, and can also produce effects that cascade 
across trophic levels [@thomas2004extinction; @edgar2010nino; @moreno2011extreme; @hoover2014resistance]. 
For example, @chiu2013impact showed that extreme floods affected 
the annual survival of a prey community, in turn influencing predators in 
aquatic as well as adjacent riparian habitats. 
@dreesen2014successive, in an experimental study, showed that 
ECEs can reduce the ability of plant communities to withstand subsequent ECEs
that occur after a short time interval.

@@

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
for both Q1 and Q2 instead of attempting mathematical 
analysis and/or using models that are more realistic, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms outlined above. 
By so doing we hope to open these potentially practically important questions to 
wider examination by other researchers, using a variety of models and perspectives. 
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula),
and our essential goal is to advance the idea that tail associations may
also be important in this context. 
Our questions are also an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.

@@

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have the same Spearman 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Spearman's $\rho$ was the same, $\rho =$ `r srho_common`,
up to sampling variance (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of $1/2$.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. 
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables. These ideas are revisited in the Discussion.

<!--FIG : schematic diagram for partial spearman correlation, adapted from BIVAN-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=6 cm]{./Results/PartialSpearmanFig.pdf}
\caption{The partial Spearman correlation, $\cor_{l_b,u_b}(u, v)$, within a band can be computed
for any band to describe how the strength of association between $u$ and $v$
varies from one part of the two distributions to another. Diagonal lines 
show two bands, the data in the right/upper band showing
stronger association than those in the left/lower band. Each band is bounded by
the lines $x+y=2l_b$ and $x+y=2u_b$ described in Methods (for values of $l_b$ and $u_b$
that depend on the band to be used). This figure was reproduced
largely following Fig. 7 of Ghosh \emph{et al.} (2019).
\label{PartialSpearmanFig}}
\end{center}
\end{figure*}

@@

@@

<!--Q1 : age-structured model results-->
For our age-structured, non-spatial model, extinction risks were higher when 
associations between the vital rates $s_e$ and $f_a$ were principally for low values
of these vital rates (left-tail association), and were lower when associations
were principally for higher values; but the effect was minor for this model (Fig. \ref{fig_extrisk_agestr}).
Though extinction risks were overall higher for the density-dependent version of our model compared
to density-independent version, extinction risk differences across tail association 
scenarios were similar for the two versions. Because the same Spearman correlation (up to sampling
variance) was used in all simulations, the differences between lines in the panels
of Fig. \ref{fig_extrisk_agestr} are attributable solely to the changes we made in the
asymmetry of tail associations, and answer Q1. Though differences were significant (confidence
bands in Fig. \ref{fig_extrisk_agestr}, which are very narrow, did not overlap),
they were also small, corresponding to only a few percentage points of additional extinction
risk after 25 years. A potential reason for this outcome is elaborated in the Discussion.

<!--Q2 : metapopulation results-->
For our metapopulation simulations, for low values of $r$ (close to $r_{min}$), left-tail
associations of environmental noises accentuated extinction risks, and right-tail associations
mitigated them; but the reverse was true for high values of $r$ (close to
$r_{bf}$). Thus 
density dependent models can either produce similar results, regarding the influence of 
tail associations on extinction risk, as the density-independent models of @Ghosh_copula, or
opposite results, depending on the nature of density dependence. 
For each model of Table \ref{tab_model_summary}, using a value of $r$ on
the low end of the range $(r_{min},r_{bf})$, extinction risk was highest
for extreme left-tail associated noise, lowest for extreme right-tail associated
noise, and was intermediate for the moderate tail association scenarios, for
both local (Fig. \ref{fig_extrisk_allmodel_localdisp}, left panels) and global 
(Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}, left panels) dispersal, and for 
essentially all the dispersal rates, $d$, that we considered. However, 
using a value of $r$ on the high end of the range $(r_{min},r_{bf})$,
extinction risks were ordered oppositely (right panels of Figs 
\ref{fig_extrisk_allmodel_localdisp} and \ref{SM-fig_extrisk_allmodel_globaldisp}). 
These results help address Q2 from the Introduction.
Intermediate values of $r$ are shown in Figs. \ref{SM-fig_ricker_risk_local_disp_scl_1} - 
\ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} for local dispersal and 
Figs. \ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6} for global dispersal among 
habitat patches.

@@

<!--summary results:-->
Our results combined with earlier results 
show that asymmetries in tail associations between environmental variables in
different locations can have a substantial influence on metapopulation extinction risk,
for density-dependent as well as for density-independent 
models. These results
extend the result of @Ghosh_copula, which was for a density-independent model. For a
density-independent model, @Ghosh_copula corroborated the intuition
that, when "harmful" environmental events 
(i.e., events which reduce the growth rates of local populations)
are correlated across space,
extinction risk of metapopulations will be accentuated compared 
to when "beneficial" events
(i.e., events which increase the growth of local populations)
are spatially correlated, even if overall correlation is the same in 
both scenarios. This was presumably because the first of these scenarios reduces the 
potential for rescue effects. For the density-dependent metapopulation 
models we considered, the same 
result held, typically, but not always, for undercompensatory model dynamics. However,
for overcompensatory dynamics, the oposite result held, again typically, but not always: 
when "harmful" events were correlated across space,
extinction risk of metapopulations was actually mitigated compared 
to when "beneficial" environmental events were 
spatially correlated by the same amount. This was probably because the correlated 
beneficial events caused populations to exceed carrying capacity, and 
to subsequently crash, across all or most habitat patches. In contrast 
to the metapopulation case, our results 
suggest that asymmetry of tail associations between vital rates
in a stage-structured, non-spatial stochastic model may be of limited 
importance for extinction risk. In our stage-structured, non-spatial model, extinction 
risk was only weakly influenced by asymmetry of tail associations
between vital rates. Our results provide suggestive, rather
than definitive evidence, here, because we considered 
only one model (but see below).

<!--linking back to ECEs-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and even specific events can now be attributed to
human-induced climate change [@ummenhofer2017extreme]. We here provide a fuller 
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places, and thereby influence metapopulation
extinction risk through the mechanisms we explored in this paper. 
If $X_i(t)$ denotes the value of an environmental variable (e.g., a summer maximum 
temperature, or winter minimum temperature, spring rainfall, etc.) 
in location $i$ in year $t$,
@ummenhofer2017extreme provide a diagram of how ECEs may be increasing in frequency 
and intensity, linked to changes in the distribution of $X_i(t)$ such as changes
in the mean, variance, or skewness of this distribution. Although changes in 
the correlation between $X_i(t)$ and $X_j(t)$ are increasingly documented, 
and may also be a consequence of climate change (e.g., @Black2018),
we are unaware of any studies examining the extent to which tail associations
between $X_i(t)$ and $X_j(t)$, and asymmetries of tail association, may be 
changing. However, documented changes in correlation between $X_i(t)$ and $X_j(t)$
may be due partly or wholly to changes in tail associations. If ECEs are becoming
more spatially extensive, as seems likely given that they are becoming
more intense [@ummenhofer2017extreme], then the conditional probability that both of
$X_i(t)$ and $X_j(t)$ are influenced by an ECE, given that one of them is so
influenced, should be increasing. This should contribute to increased association
of the two variables in one of their tails (which tail depends on the type of 
ECE and what the variable $X$ is). A study of changes through time in 
tail associations of weather variables would be straightforward and 
potentially important given the mechanisms we have explored.
<!--***DAN: Shyamolina, another aspect of an exploration you may do on the 
the geography of tail association may be to look at how tail association 
may be changing. This relates to ECEs, so may help better justify the work
to a funder. You had proposed to look at geography of tail association in 
NDVI, and that makes good sense. You could also look at how tail association
is changing in NDVI (the records are about 35 years long, so you could just
about do this). Even better (because data are better), you could get weather
station records (lots of these can be downloaded, over 1000 stations
have been operating for over a century continuously in North America) and 
look at how tail associations are changing in those data.-->
